---
title: "R Objects"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(nycflights13)
library(openintro)
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

In R there are many different objects that are used for various problems
and circumstances. In this tutorial we show the most relevant ones that will
be very useful during the course. 

Generally an object is a variable that stores some data. Depending on the
data that we have we should use different objects. The most simple object is
just a variable that stores a number, then we can extent this to a vector of numbers.
We also describe very useful and general objects called lists and finish by 
discussion of data frames that are special kind of lists and used most frequently
in data analysis.

## Atomic Vectors

R lets you save data by storing it inside an R object.
What is an object? Just a name that you can use to call up stored data.

For example let's create a variable `x` and store 1 in it:
```{r echo=TRUE, eval=TRUE}
x <- 1
x
```
The symbol `<-` means assignment of 1 to `x`. To see what data any variable stores
just type the name of the variable (`x` in our case) and press enter (for now please ignore `[1]` at the beginning of the output) . 

In many cases we want to group some values of data together in one object (called a vector),
we can easily do that in R using function c:
```{r echo=TRUE, eval=TRUE}
die <- c(1,2,3,4,5,6)
die
```
Now the object `die` contains 6 values. We can check whether an object is a vector or not
using `is.vector` function:
```{r echo=TRUE, eval=TRUE}
is.vector(die)
```
For an object that is a vector it is frequently useful to know how many elements are in that vector,
we can get it using `length` function:
```{r echo=TRUE, eval=TRUE}
length(die)
```
Note that when we store only one number to a variable, R considers it as a vector
of length 1:
```{r echo=TRUE, eval=TRUE}
x <- 1
is.vector(x); length(x)
```
To access a value of a vector at particular position we use `[]`, with value
of the position inside the brackets. Here is an example, we extract the third element
of a vector:
```{r echo=TRUE, eval=TRUE}
myvec <- c(5,3,6,1,4,2)
myvec[3]
```
We can input a vector of positions to extract more than one value from a vector:
```{r echo=TRUE, eval=TRUE}
myvec <- c(5,3,6,1,4,2)
myvec[c(1,3,5)]
```

### Exercises

A. Save number 2 to variable `y`, add 3 to `y` and see the result:
```{r, ex1.1, exercise=TRUE}
y <-
```
```{r ex1.1-hint-1}
y <- 3
```
```{r ex1.1-solution}
y <- 3
y + 2
```

B. Save a vector with the name `myvec` that contains the values 1,3,2,4. Check
whether it is a vector and find the length of it:
```{r ex1.2, exercise=TRUE}
myvec <-
```
```{r ex1.2-hint-1}
myvec <- c(1,3,2,4)
```
```{r ex1.2-solution}
myvec <- c(1,3,2,4)
is.vector(myvec); length(myvec)
```

C. Consider the following vector:
```{r echo=FALSE, eval=TRUE, cache=TRUE}
set.seed(Sys.time())
x <- sample(1:10,5,replace=FALSE)
x
```
Use R to answer the following questions:
```{r ex1.3, exercise=TRUE}
```
```{r ex1.3-question, echo=FALSE}
l <- length(x)
quiz(caption = "",
     question("What is the length of the given vector?",
              answer(sprintf("%s",l+1)),
              answer(sprintf("%s",l), correct = TRUE),
              answer(sprintf("%s",l-1)),
              answer(sprintf("%s",l-2)),
              random_answer_order = TRUE,
              allow_retry = FALSE
     ),
     question("What is the value the vector stores on SECOND position?",
              answer(sprintf("%s",x[2]),correct=TRUE),
              answer(sprintf("%s",x[1])),
              answer(sprintf("%s",x[3])),
              answer(sprintf("%s",x[4])),
              random_answer_order = TRUE,
              allow_retry = FALSE
     )
)
```

## Type of vectors

Each atomic vector can only store one type of data. You can save different types of data in R by using
different types of atomic vectors.
R recognizes six basic types of atomic vectors: doubles, integers, characters, logicals, complex, and raw.

We will not consider complex and raw types here.
In previous section where we saved some numbers to a vector, R automatically
stored them as doubles (or just numeric) type.
For integer vectors we must include a capital L with input, and for character vectors we have to surround an input by
quotation marks:
```{r echo=TRUE, cache=TRUE}
num <- c(2,3)
int <- c(2L, 3L)
coin <- c("Tail","Head")
num
int
coin
```
Now object `int` is a vector that contains two integers (2 and 3) while
the object `coin` is a vector of two characters/words ("Tail" and "Head").
We can use `typeof` function to get the type of a vector:
```{r echo=TRUE, cache=TRUE}
typeof(num); typeof(int); typeof(coin)
```
The last type that will discuss is logicals. Logical vectors store TRUEs and FALSEs, R’s form of Boolean data. 
TRUE or FALSE in capital letters (without quotation marks) will be treated as logical data:
```{r echo=TRUE, eval=TRUE}
logic <- c(TRUE,FALSE,FALSE,TRUE)
logic
```
Logicals are very helpful for comparisons, for example if we are given a vector
of random numbers between 1 and 10 we can check which ones are greater than 5:
```{r echo=TRUE, eval=TRUE, cache=TRUE}
x <- c(6,1,4,8,4,6,7,3)
comp <- x>5
comp
```
Moreover R can consider a logical vector as a vector of 0s and 1s (corresponding
to FALSEs and TRUEs respectively). So if we need the number of elements in vector `x` that
are larger than 5, we can just sum all elements in vector `comp`:
```{r echo=TRUE, eval=TRUE}
sum(comp)
```

### Exercises

A. Create a vector `int` that stores integers 3,5,1,2, and check the type of this vector:
```{r ex2.1, exercise=TRUE}
int <-
```
```{r ex2.1-hint-1}
int <- c(3L,5L,1L,2L)
```
```{r ex2.1-solution}
int <- c(3L,5L,1L,2L)
typeof(int)
```

B. Given the following vector:
```{r echo=FALSE, eval=TRUE, cache=TRUE}
set.seed(Sys.time())
x <- sample(1:10,7,replace=FALSE)
x
```
```{r ex2.2, exercise=TRUE}
```
```{r ex2.2-question, echo=FALSE}
x1 <- sum(x<4)
x2 <- sum(x>6)
quiz(caption = "Using R find the following:",
     question("How many elements in the vector are smaller than 4?",
              answer(sprintf("%s",x1+1)),
              answer(sprintf("%s",x1), correct = TRUE),
              answer(sprintf("%s",x1+2)),
              answer(sprintf("%s",x1+3)),
              random_answer_order = TRUE,
              allow_retry = FALSE
     ),
     question("How many elements in the vector are greater than 6?",
              answer(sprintf("%s",x2),correct=TRUE),
              answer(sprintf("%s",x2+1)),
              answer(sprintf("%s",x2+2)),
              answer(sprintf("%s",x2+3)),
              random_answer_order = TRUE,
              allow_retry = FALSE
     )
)
```

C. We can assign names to vector elements using `names` function, for example:
```{r eval=TRUE, echo=TRUE}
x <- c(0,1)
name <- c("zero","one")
names(x) <- name
x
```
Following the example construct a vector `die` with values 1,2,3,4,5 and 6 with 
names "one","two’,,"six" and see the resulting vector.
```{r ex2.3, exercise=TRUE}
die <- c(
```
```{r ex2.3-hint-1}
die <- c(1,2,3,4,5,6)
```
```{r ex2.3-hint-2}
die <- c(1,2,3,4,5,6)
name <- c("one","two","three","four","five","six")
```
```{r ex2.3-solution}
die <- c(1,2,3,4,5,6)
name <- c("one","two","three","four","five","six")
names(die) <- name
die
```

## Matrices

Matrices are just atomic vectors that are organized into rectangular array.
Similar to vectors, elements in a matrix must be of the same type (doubles, integers, and so on.).
Number of rows `nrow` and number of columns `ncol` characterize matrices. As an example
consider the following matrix:
```{r echo=TRUE, eval=TRUE}
mat <- matrix(c(1,2,3,4,5,6),nrow=2)
mat
```
Note that we do not have to specify the number of columns since the length of vector (which is 6)
implies that the matrix must have 3 columns. Another important observation is that the matrix is
'filled' column by column (by default), if we need by rows we have to add `byrow=TRUE`:
```{r echo=TRUE, eval=TRUE, cache=TRUE}
mat <- matrix(c(1,2,3,4,5,6),nrow=2, byrow=TRUE)
mat
```
Using functions `rownames` and `colnames` we can give names to rows and columns of a matrix
respectively:
```{r echo=TRUE, eval=TRUE, cache=TRUE}
rownames(mat) <- c("R1","R2") 
colnames(mat) <- c("C1","C2","C3")
mat
```
Sometimes it might be of interest to find the dimension of a matrix, we can 
easily get it with `dim` function:
```{r echo=TRUE, eval=TRUE}
dim(mat)
```
The first number represents number of rows, second, number of columns.

To access some particular entries in a matrix we use a pair of square brackets `[,]` with the
first and second numbers indicating position of row and position of column respectively:
```{r echo=TRUE, eval=TRUE}
mat[1,2]; mat[2,3]
```
Note that if a matrix has column and row names we can input specific names instead of numbers,
we get exactly the same results with the next code:
```{r echo=TRUE, eval=TRUE}
mat["R1","C2"]; mat["R2","C3"]
```
We can extract entire row if we leave the second entry in squared brackets empty,
similarly we get the entire column by leaving the first entry blank:
```{r echo=TRUE, eval=TRUE}
mat[2,]; mat[,3]
```
In the first example we get second row (as a vector) followed by a third column (also as a vector).
Using similar `c()` function we can extract several rows and/or columns. For
example the next code retrieves two entries from the second row and from columns 1 and 3:
```{r echo=TRUE, eval=TRUE}
mat[2,c(1,3)]
``` 

### Exercises

A. Create a 3 by 3 matrix `mat` with 5s on the diagonal and 0s everywhere else and see the resulting matrix:
```{r ex4.1, exercise=TRUE} 
mat <- 
```
```{r ex4.1-hint-1} 
mat <- matrix(c(5,0,0,0,5,0,0,0,5))
```
```{r ex4.1-solution} 
mat <- matrix(c(5,0,0,0,5,0,0,0,5),ncol=3)
mat
```

B. Given the following matrix called `mymat':
```{r echo=FALSE, eval=TRUE, cache=TRUE}
set.seed(Sys.time())
mymat <- matrix(sample(1:20,10,replace=FALSE),ncol=5)
mymat
```
```{r ex4.2, exercise=TRUE}
mymat
```
```{r ex4.2-question, echo=FALSE}
quiz(caption = "Using R find the following:",
     question("What is the dimension of 'mymat'?",
              answer("(1,6)"),
              answer("(2,5)", correct = TRUE),
              answer("(3,4)"),
              answer("(2,6)"),
              random_answer_order = TRUE,
              allow_retry = FALSE
     ),
     question("What is the value of 'mymat' on second row and third column?",
              answer(sprintf("%s",mymat[2,3]),correct=TRUE),
              answer(sprintf("%s",mymat[2,4])),
              answer(sprintf("%s",mymat[1,3])),
              answer(sprintf("%s",mymat[1,2])),
              random_answer_order = TRUE,
              allow_retry = FALSE
     )
)
```

## Lists

Lists are like atomic vectors because they group data into a one-dimensional set.
Unlike vectors, lists can store different types of variables and different objects.
Lists group together R objects, such as atomic vectors, matrices and other lists.
For example, you can make a list that contains a numeric vector of length 6 in its first element, a
character vector of length 1 in its second element, a 2 by 3 matrix in its third element and a list of length 2 in its fourth element.
```{r eval=TRUE, echo=TRUE}
mylist <- list( c(1,2,3,4,5,6), "Red", matrix(c(1,2,3,4,5,6),ncol=3), list(TRUE,1))
mylist
```
As with atomic vectors we can use function `names` to give names for each element in the list:
```{r eval=TRUE, echo=TRUE,cache=TRUE}
names(mylist) <- c("Vector","Color","Matrix","List")
mylist
```
To access a particular element in the list we can use double square brackets `[[]]` with position of the 
element inside the brackets:
```{r eval=TRUE, echo=TRUE}
mylist[[1]]; mylist[[2]]
```
If entries in the array have names, we can use `$` sign followed by the name of a variable (no quotations) instead of square brackets:
```{r eval=TRUE, echo=TRUE}
mylist$Vector; mylist$Color
```
Once we have a variable from the list that we are interested in, we can apply usual rules for manipulating this object. For example
if we need to extract an entry on first row and third column of a matrix which is in `mylist` we can do the following:
```{r eval=TRUE, echo=TRUE}
mylist[[3]][1,3]
``` 

### Exercises

A. Create a list called `list_1` that has 3 elements: logical vector (TRUE,FALSE), character "GREEN" and numerical vector (3,2,1).
Next give name for each element: "Logic", "Color", "Vector" and see the result:
```{r ex5.1, exercise=TRUE} 
list_1 <- 
```
```{r ex5.1-hint-1} 
list_1 <- list(c(TRUE,FALSE),
```
```{r ex5.1-hint-2} 
list_1 <- list(c(TRUE,FALSE), "GREEN",
```
```{r ex5.1-hint-3} 
list_1 <- list(c(TRUE,FALSE), "GREEN", c(3,2,1))
```
```{r ex5.1-hint-4} 
list_1 <- list(c(TRUE,FALSE), "GREEN", c(3,2,1))
names(list_1) <- c("Logic",
```
```{r ex5.1-hint-5} 
list_1 <- list(c(TRUE,FALSE), "GREEN", c(3,2,1))
names(list_1) <- c("Logic","Color",
```
```{r ex5.1-solution} 
list_1 <- list(c(TRUE,FALSE), "GREEN", c(3,2,1))
names(list_1) <- c("Logic","Color","Vector")
list_1
```

B. Given the following list called `mylist':
```{r echo=FALSE, eval=TRUE, cache=TRUE}
set.seed(Sys.time())
mylist <- list(NA)
mylist[[1]] <- matrix(sample(1:20,10,replace=FALSE),ncol=5)
mylist[[2]] <- sample(1:20,4,replace=FALSE)
names(mylist) <- c("Mat","Vec")
mylist
```
```{r ex5.2, exercise=TRUE}
mylist
```
```{r ex5.2-question, echo=FALSE}
quiz(caption = "Using R find the following:",
     question("What is dimension of a matrix 'Mat' in this list?",
              answer("(1,6)"),
              answer("(2,5)", correct = TRUE),
              answer("(3,4)"),
              answer("(2,6)"),
              random_answer_order = TRUE,
              allow_retry = FALSE
     ),
     question("What is the entry of matrix 'Mat' on second row and fifth column?",
              answer(sprintf("%s",mylist[[1]][2,5]),correct=TRUE),
              answer(sprintf("%s",mylist[[1]][2,3])),
              answer(sprintf("%s",mylist[[1]][1,3])),
              answer(sprintf("%s",mylist[[1]][1,4])),
              random_answer_order = TRUE,
              allow_retry = FALSE
     ),
     question("What is the length of a vector 'Vec' in this list?",
              answer("5"),
              answer("4", correct = TRUE),
              answer("3"),
              answer("6"),
              random_answer_order = TRUE,
              allow_retry = FALSE
     ),
    question("What is the value on the third position of a vector `Vec'?",
              answer(sprintf("%s",mylist[[2]][3]),correct=TRUE),
              answer(sprintf("%s",mylist[[2]][4])),
              answer(sprintf("%s",mylist[[2]][2])),
              answer(sprintf("%s",mylist[[2]][1])),
              random_answer_order = TRUE,
              allow_retry = FALSE
     )
)
```

## Data Frames

Data frames are probably the most useful data structure that is used in data analysis. A data frame is a list
of atomic vectors of the same length (but of different types). It is equivalent to the Excel spreadsheet with
columns and rows correspond to variables and observations respectively. Note that each column (or vector in our case) of a data frame
can contain a different type of data but within a column every cell must be of the same type. Lets create our first
data frame using function `data.frame` consisting of 4 observations and two variables. First variable is a student number, second is a student's name:
```{r echo=TRUE, eval=TRUE}
num <- c(1,2,3,4)
name <- c("Nadia","Shiyi","Yizhe","Wei")
mydat <- data.frame(student_num = num, student_name = name)
mydat
```
Note that each vector turned into a column and that the names of the variables are `student_num` and `student_name`. These
two variables (vectors) have the same number of observations (which is 4) but are of different types, one is numeric
another is a factor (by default R changes characters to factors). We can check whether an object is a data frame using 
`is.data.frame` function:
```{r eval=T,echo=T}
is.data.frame(mydat)
```
Another useful function is `dim` which for a data frame outputs two numbers, the number of rows and number
of columns:
```{r eval=T,echo=T}
dim(mydat)
```
Usually we will upload data sets to R from some
external sources, however R has some build-in data frames for analysis. For example data set `mtcars` contains
performances of 32 automobiles, to see the first 6 rows of the data frame we use function `head`:
```{r echo=TRUE, eval=TRUE}
head(mtcars)
```
To extract a variable from a data frame we can use squared brackets `[,]` where second number corresponds
to position of the column in the data frame (similar to matrices) or we can use `$` sign followed by the name
of the variable (as with lists). For example lets extract `hp` variable from the `mtcars` data frame and 
find its mean using `mean` function:
```{r echo=TRUE, eval=TRUE}
hp <- mtcars[,4]; mean(hp)
hp <- mtcars$hp;  mean(hp)
``` 
If we need for example the value of `hp` variable only from the second row, again we can either use
square brackets with the first number equals to 2 or first use `$` sign to extract the vector and then 
get the value on the second position:
```{r echo=TRUE, eval=TRUE}
mtcars[2,4]; mtcars$hp[2]
```
Generally the use of `[,]` is very versatile and with it we can extract any combinations of rows (observations)
and columns (variables). Frequently we need to analyze (or extract) rows that correspond to some logical criteria,
for example, suppose we need to extract all rows (and all columns) from `mtcars` for which `hp` is greater than 200, we can easily
do that with the combination of `[,]` and `$`:
```{r echo=TRUE, eval=TRUE}
mtcars[mtcars$hp>200,]
```
Note that we left second number after comma in square brackets blank since we need all variables. Now lets
extract all observations for which `mpg` is greater than 16 AND `cyl` is 8:
```{r echo=TRUE, eval=TRUE}
mtcars[mtcars$mpg>16 & mtcars$cyl==8,]
```
Two important observations from the last example are that logical “AND” is represented by `&` in R and for
checking equality in logical statements we have to enter double equal sign `==`. If instead of “AND” statement
we had “OR”, the code would remain the same except that `&` must be changed to `|`. 

### Exercises

A. Create a data frame called `mydata` that has 3 variables: Die = (1,2,3,4,5,6), Color = ("red","red","green","green","blue","blue")
and Logic = (TRUE,TRUE,TRUE,TRUE,FALSE,FALSE). Check whether it is a data frame, find its dimension and see the result:
```{r ex6.1, exercise=TRUE} 
mydata <- data.frame( 
```
```{r ex6.1-hint-1} 
mydata <- data.frame(Die = c(1:6),
```
```{r ex6.1-hint-2} 
mydata <- data.frame(Die = c(1:6), Color = c("red","red","green","green","blue","blue"), 
```
```{r ex6.1-hint-3} 
mydata <- data.frame(Die = c(1:6), Color = c("red","red","green","green","blue","blue"), Logic = c(TRUE,TRUE,TRUE,TRUE,FALSE,FALSE) )
```
```{r ex6.1-hint-4} 
mydata <- data.frame(Die = c(1:6), Color = c("red","red","green","green","blue","blue"), Logic = c(TRUE,TRUE,TRUE,TRUE,FALSE,FALSE) )
is.data.frame(mydata)
```
```{r ex6.1-solution} 
mydata <- data.frame(Die = c(1:6), Color = c("red","red","green","green","blue","blue"), Logic = c(TRUE,TRUE,TRUE,TRUE,FALSE,FALSE) )
is.data.frame(mydata)
dim(mydata)
mydata
```

B. Given the following data frame called `data':
```{r echo=FALSE, eval=TRUE, cache=TRUE}
set.seed(Sys.time())
vec1 <- sample(1:100,10,replace=FALSE)
vec2 <- sample(1:20,10,replace=FALSE)
vec3 <- c("a","b",sample(c("a","b"),8,replace=TRUE))

data <- data.frame( Var_1=vec1, Var_2=vec2, Var_3=vec3)
data
```
```{r ex6.2, exercise=TRUE}
data
```
```{r ex6.2-hint-1}
sum(data$Var_3=="a")
```
```{r ex6.2-hint-2}
mean(data[data$Var_3=="b",2])
```

```{r ex6.2-question, echo=FALSE}
quiz(caption = "Using R find the following:",
     question("Find the mean of Variable 1?",
              answer(sprintf("%s",round(mean(data$Var_1),2)),correct=TRUE),
              answer(sprintf("%s",round(mean(data$Var_2),2))),
              answer(sprintf("%s",round(mean(data$Var_1)+6,2))),
              answer(sprintf("%s",round(mean(data$Var_1)-2,2))),
              random_answer_order = TRUE,
              allow_retry = FALSE
     ),
     question("What is the value on fifth row and second column of this data frame?",
              answer(sprintf("%s",data[5,2]),correct=TRUE),
              answer(sprintf("%s",data[4,2])),
              answer(sprintf("%s",data[6,2])),
              answer(sprintf("%s",data[7,2])),
              random_answer_order = TRUE,
              allow_retry = FALSE
     ),
     question("How many rows are there with 'a' in Variable 3?",
              answer(sprintf("%s",sum(data$Var_3=="a")),correct=TRUE),
              answer(sprintf("%s",sum(data$Var_3=="a")-1)),
              answer(sprintf("%s",sum(data$Var_3=="a")+1)),
              answer(sprintf("%s",sum(data$Var_3=="a")+2)),
              random_answer_order = TRUE,
              allow_retry = FALSE
     ),
    question("What is the mean of Variable 2 when Variable 3 is 'b'?",
              answer(sprintf("%s",round(mean(data[data$Var_3=="b",2]),2)),correct=TRUE),
              answer(sprintf("%s",round(mean(data[data$Var_3=="b",1]),2))),
              answer(sprintf("%s",round(mean(data[data$Var_3=="b",2])+2,2))),
              answer(sprintf("%s",round(mean(data[data$Var_3=="b",2])-3,2))),
              random_answer_order = TRUE,
              allow_retry = FALSE
     )
)
```